---
title: "Total and estimated number of US respondents that had a doctorate degree as highest level of attainment"
author: 
  - Chris Yong Hong Sen
  - Xuanang Ren
  - Caichen Sun
  - Irene Liu
  - Mingxuan Liu
thanks: "Code and data are available at: [https://github.com/Monoji77/USA_data](https://github.com/Monoji77/USA_data)."
date: today
date-format: long
format: pdf
number-sections: true
bibliography: references.bib
bibliographystyle: apa
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
usa_data <- read_csv("../data/01-raw_data/usa_00001.csv")
set.seed(637)
```


# Introduction {#sec-intro}
We will be analysing how many respondents were there in each state (STATEICP) that had a doctoral degree as their highest educational attainment (EDUC). Particularly, we aim to use Laplace ratio estimators to support our findings.

The remainder of this paper is structured as follows. @sec-data will contain information about data retrieval and data manipulation using @citeR. 




# Data {#sec-data}

## Data Retrieval
Our dataset was obtained from USA Ipums, which is an online US census data for social, economic and health research resource [@citeUSAIPUMS]. Following the instructions outlined by USA Ipums, we are able to select 2022 American Census Dataset, where we subset the state ($STATEICP$) and educational attainment ($EDUCD$) of the individual respondent. These variables are relevant to our research question introduced in @sec-intro.

## Ratio Estimators Approach Overview
As for data manipulation, we use tidyverse library from R and we can obtain the total number of respondents in the state of california [@citeR]. Furthermore, we can also obtain the total number of respondents in California with a doctorate degree by finding EDUCD value of 116. We were able to find this value in the documentation provided by USA Ipums [@citeUSAIPUMSDocumentation]. Finally the laplace ratio can be calculated as...


\text{californian\_doctorates\_total\_respondents} / \text{californian\_total\_respondents}

We then use this value obtained and multiply it by the number of respondents in other states to determine an estimate number of respondents in that state with a doctorate degree. @tbl-states show us the top few entries of actual number of doctorates per US state and estimated number of doctorates per US state using the aforementioned laplace ratio. Note that the estimated values could differ since the ratio between doctorate respondents and total respondents in california could be different from that of another state. One reason this could be is that there are more universities in some states compared to california (vice versa for states with lesser universities). There are other socio-demographic related reasons as well such as gender and family's household income. This is reflected in @tbl-states where we can observe how the difference between estimated and actual total number of doctorate respondents from the first observation only differs by 0.8% while that of the second observation had a larger difference of 42%.


```{r raw code to obtain results}
#| label: tbl-states
#| tbl-cap: states with doctoral as educational attainment
#| echo: false

######################
# goal 1: obtain laplace ratio using california data
#
######################

#californian ratio
california_doctorate <- usa_data |>
  select(STATEICP, EDUCD) |>
  filter(EDUCD == 116) |> # 116 refers to Doctoral degree 
  group_by(STATEICP) |>
  count(name = 'no_of_respondants_doctorate') |>
  filter(STATEICP == 71) |>
  pull(no_of_respondants_doctorate)

california_total <- usa_data |>
  select(STATEICP, EDUCD) |> # 116 refers to Doctoral degree 
  group_by(STATEICP) |>
  count(name = 'no_of_respondants') |>
  filter(STATEICP == 71) |>
  pull(no_of_respondants)

laplace_ratio <- california_doctorate / california_total


####################
# goal 2: estimate total number of doctorates using laplace ratio
#
####################

# specific to Doctorates per state
no_of_respondants_doctorate <- usa_data |>
  select(STATEICP, EDUCD) |>
  filter(EDUCD == 116) |> # 116 refers to Doctoral degree 
  group_by(STATEICP) |>
  count(name = 'no_of_respondants_doctorate')

#no_of_respondants_doctorate

# total respondents per state
no_of_respondants <- usa_data |>
  select(STATEICP, EDUCD) |>
  group_by(STATEICP) |>
  count(name = 'no_of_respondants')
#no_of_respondants

# join both tables
no_of_respondants_doctorate |>
  inner_join(no_of_respondants,
             by= join_by(STATEICP)) |>
  mutate(estimate_no_of_respondants_doctorate = 
           round(laplace_ratio * no_of_respondants, 0),
         .after=no_of_respondants_doctorate) |>
  ungroup() |>
  select(`state ID` = STATEICP, 
         `actual doctorate count` = no_of_respondants_doctorate, 
         `estimated doctorate count`= estimate_no_of_respondants_doctorate)
```





# References


